### 2021-09-03 revision/formerPapers/ens-rev/reviewsections/ancey2008.tex
added phd thesis.

We can consider the flow of probability up and down the ladder of states ($n=0,1,2,\dots$) which is induced by these transitions, entrainment (including the collective entrainment term), deposition, immigration, and emigration, using the formalism of more general Markov birth-death processes \citep{Cox1965, Pielou1977}. 
By the standard way of developing master equations \citep[e.g.][]{Cox1965}, we find an infinite system of equations for the probabilities of finding $n$ moving particles within the control volume: 
\begin{multline} P(n,t+\delta t) = \alpha \delta t (n+1)P(n+1,t) \\+  [\lambda + (n-1)\mu]\delta t P(n-1,t) \\ + [1-\delta t[\lambda + n\alpha + n\mu]]P(n,t). \end{multline}
Here $\alpha = \nu + \sigma$ summarizes the contributions of emigration and deposition, which both act to decrease the number of particles $n$, and $\lambda = \lambda_0 + \lambda_1$ summarizes the contributions of immigration and (individual) entrainment, which both act to increase the number of particles $n$.

@@

In effect, the probability generating function exchanges the discrete variable $n$ for a continous variable $z$. 
This first order partial differential equation for $G$ can be solved by the method of characteristics \citep{Cox1965, Garabedian1964}. 
If there are initially $N_0$ moving particles within the control volume, the solution is 
\be G(z,t) + \Big(\frac{\alpha-\mu}{(K-1)\mu z + \alpha - K \mu}\Big)^{n+\lambda/\mu}\Big( \frac{(K\alpha-\mu)z + \alpha(1-K)}{\alpha-\mu}\Big)^n. \label{eq:generator}\ee
The factor $K$ is the autocorrelation function $K(t) = \exp(-t(\alpha-\mu)).$

@@

As discussed, the negative binomial distribution supports wide fluctuations, controlled by the collective entrainment parameter $\mu$. 
The mean number of particles in the control volume, taken over the distribution \ref{eq:negbin}, is 
\be \bra n \ket = \sum_{n=0}^\infty n P(n) = \frac{\lambda}{\alpha-\mu}. \label{eq:anceymean}\ee
Likewise the variance is 
\be \bra (\delta n)^2 \ket = \sum_{n=0}^\infty (n-\bra n\ket)^2 P(n) = \frac{ \lambda \alpha}{\alpha-\mu^2},\ee
which means their ratio is 
\be \frac{\bra (\delta n)^2}{\bra n\ket} = \frac{1}{1-\mu/\alpha}.\ee
It is interesting to compare this to equation \ref{eq:2006flucts} bounding the magnitude of fluctuations in the \citet{Ancey2006} model. 
In contrast, within the \citet{Ancey2008} model, the variance must exceed the mean, because as discussed, in steady state $\alpha>\mu$. 
The magnitude of fluctuations can grow arbitrarily with the collective entrainment parameter $\mu$. 
This is a coherent conclusion, since collective entrainment was introduced in order to represent the collective effects of turbulent fluctuations, collision-induced entrainment, and granular avalanches; and with the expressed intent of generating realistically wide bedload fluctuations, exceeding those of the \citet{Ancey2006} model.

Armed with the probability distribution for the number of particles in the control volume, equation \ref{eq:negbin}, and the link between control volume and surface statistics derived in section \ref{sec:fluxdef}, equation \ref{eq:fluxy}, we can derive the probability distribution of the bedload flux in a similar way as equation \ref{eq:anceygauss}.
The number of particles in the control volume has been implicitly assumed to be infinite, so there is no need or means to take a large $N$ limit.

Using the law for transformation of probabilities gives 
\be P(q_s) \approx \frac{\Gamma(r+\frac{Lw}{\nu_p u_s} q_s)}{\Gamma(r)\Gamma(1+\frac{Lw}{\nu_p u_s} q_s)}p^r(1-p)^{\frac{Lw}{\nu_p u_s} q_s}\ee
as the probability distribution of the bedload flux $q_s$.

Now we link back to the \citet{Einstein1950} theory. 
Einstein considered neither immigration ($\lambda_1$), emigration ($\nu$), nor collective entrainment ($\mu$). 
When all of these parameters are set to zero, $\mu = \lambda_1 = \nu = 0$, the parameters $r$ and $p$ of the distribution \ref{eq:negbin0} become $r \rightarrow \infty$ and $p=0$. 
In this limit the distribution \ref{eq:negbin0} tends to a Poisson distribution.

There are two ways to see this. 
The first way is to take the appropriate limits ($t \rightarrow \infty$, $\mu=lambda_1=\nu= 0$)  into the generating function \ref{eq:generator} gives 
\be G(z) = e^{-\lambda_0(z-1)/\sigma},\ee
which inverts to 
\be P(n) = \frac{(\lambda_0/\sigma)^n}{n!}e^{-\lambda_0/\sigma},\label{eq:anceypoisson}\ee
the Poisson distribution.  
The second way is to take limits of the negative binomial distribution equation \ref{eq:negbin0} directly, although this is somewhat tricky.
It involves reparameterizing the distribution by its mean value equation \ref{eq:anceymean} rather than $r$, and then taking limits, using the limit formula $e^{-x} = \lim_{s \rightarrow \infty} (1-x/s)^s$.

In either case, arriving at this Poisson distribution for $n$ in the absence of collective entrainment, we can compute the mean bedload flux using equation \ref{eq:fluxy}: 
\be \bra q_s \ket = \frac{\nu_p u_s}{L w} \sum_{n=0}^\infty n \frac{(\lambda_0/\sigma)^n}{n!}e^{-\lambda_0/\sigma} = \frac{\nu_p u_s}{L w} \frac{\lambda_0}{\sigma}. \ee
To make full contact with the Einstein-like formula of Yalin, equation \ref{eq:yalinflux}, one further step is needed.

In the \citet{Ancey2008} collective entrainment model, there was no limitation on the number of particles within the control volume. 
In the notation of section \ref{sec:ancey2006}, this is equivalent to $N \rightarrow \infty$. 
In \citet{Ancey2006} the fraction of time spent in motion, $\xi$, referred to a single particle. 
We can show in the model under consideration that $\lambda_0/\sigma$ also represents a fraction of time spent in motion, although this is more difficult than it was in the arguments leading to the analogous conclusion for $\xi$.

Similar to equation \ref{eq:}, one can set up a master equation to compute the distribution of time time periods that the number $n$ remains fixed while no transitions except deposition occur.

 that particles spend in motion and to find the mean period spent in motion is $\lambda_0$.

- this model assumes infinite sediment availability. there is no limit to n. This leads to turowski
- the flux is defined in a control volume, emigration provides a pathway to defining the flux at a point. This is Heyman 2013, and Ma 2014. 
- imagine connecting cells, this is ancey2014, which leads into Heyman 2014

\begin{comment}

When the collective entrainment process is turned off, $\mu\rightarrow 0$, the Poisson distribution for the number of moving particles within the control volume is recovered, reproducing the limiting result of \citet{Ancey2006}, which has links to \citet{Einstein1950}. 
We will consider this limit of no collective entrainment. 
$\mu\rightarrow 0$, implies $r \rightarrow \infty$ and $p \rightarrow 1$. 
In this limit, the generating function G(z) becomes 
\be G(z) = \exp(-\frac{\lambda}{\alpha}[z-1]).\ee
Taylor expanding this generating function around $z=0$ generates the hirearchy of probabilities $\pi_n(t)$ in the limit of no collective entrainment:
\be \pi_n = \frac{(\lambda/\alpha)^n}{n!}e^{-\lambda/\alpha}.\ee
Therefore the number of particles in the control volume is Poisson distributed, in accord with the large $L$ limit of the \citet{Ancey2006} model in the previous section \ref{eq:poissonflux}. 
To complete the connection we turn off immigration ($\lambda_0=0$) and emigration ($\nu = 0$) as well, so that the distribution of the number of particles in the control volume reproduces the \citet{Ancey2006} result cleanly: 
\be \pi_n = \frac{(\lambda_1/\sigma)^n}{n!} e^{-\lambda_1/\sigma},\ee
which is the same as equation \ref{eq:poissonflux}. 
We note that immigration ($\lambda_0$ -- to move in) and emigration ($\nu$ -- to move out) do not fundamentally alter the form of the distribution.
Only collective entrainment ($\mu$) does this.

Now we will analyze the probability distribution of bedload flux and the magnitude of its fluctuations. 
Collective entrainment develops a negative binomial distribution for the number of active particles, and in the limit of a large control volume ($L \rightarrow \infty$), the probability distribution for the number of active particles links to the probability distribution of the bedload flux via the relationship \ref{eq:fluxy} of the previous section. 
As a reminder, this relationship was derived using an artificial device. 
To develop an ensemble of particles intersecting a flow cross section $S$ with various positions $\textbf{x}$ and velocities $\textbf{u}_p$, we traded a frequentist interpretation of a conditional probability $P[\textbf{u}_p|\textbf{x}]$ for an ensemble one.
This was only valid in the limit of an infinitely long control volume $L \rightarrow \infty$.

The \citet{Ancey2008} model reveals another way to compute the flux which does not depend on this device. 
Considering the downstream boundary of the control volume as the surface $S$ in the formula \ref{eq:anceyflux}, we can count the number of emigration events across this surface per unit time.  
There are four types of transitions in the \citet{Ancey2008} model, and these occur at random intervals. 
Counting the frequency of only one of these transitions is a non-trivial problem, and it relies on advanced techniques in stochastic mathematics. 
This counting problem was solved by \citet{Ma2014}.
They computed the bedload flux without the need for an exchange of averaging or an artificial control volume limit ($L\rightarrow \infty$). 
We will consider this work later in the review.

For now, consider the approach of the previous section. 
The bedload flux is $q_s \approx \frac{\nu_p u_s}{L} n$, in the limit of a large control volume $L\rightarrow \infty$ and when all particles in motion approach the same velocity $u_s$.
The mean and variance of the negative binomial distribution are 
\be \bra n \ket = \frac{\lambda}{\alpha-\mu} \ee
and 
\be \bra (\delta n)^2 \ket = \frac{\lambda \alpha }{\alpha-\mu}.\ee
Because of collective entrainment, characterized by $\mu$, the variance can exceed the mean in this model: 
\be \frac{\bra (\delta n)^2 \ket}{\bra n \ket } = \frac{1}{1-\mu/\alpha}, \ee
because $1/(1-\mu/\alpha) > 1$ if $\mu > 0 $.

By these relationships, the mean flux becomes 
\be  \bra q_s \ket = \frac{\nu_p u_s}{L} \ee,
and the magnitude of bedload fluctuations is

The Ancey et al 2008 inclusion of collective entrainment develops a Negative Binomial distribution of the bedload flux. 
Expressing the bedload flux as the number of moving particles within the control volume times their average velocity $u_p$ makes a Negative binomial distribution for bedload flux within the control volume. The mean bedload flux is $ \frac{\lambda}{\alpha-\mu} u_p $ and the variance of the flux is $ \frac{\lambda \alpha}{(\alpha-\mu)^2} u_p^2$: the fluctuations are much wider in the negative binomial distribution, meaning the inclusion of collective entrainment can model the large fluctuations observed in the \citet{Ancey2008} experiments.

One way to view the bedload flux we just presented: it is the number of moving particles times their velocity. 
However, since this model explicitly includes downstream emigration from the control volume, there is another way to define the flux: it is the number of emigration events within a unit of time. 
Counting the number of emigration events expressed from this model within a unit time is not easy. 
\citet{Ma2014b} explored this approach to calculating the bedload flux, and I will review this in a subsequent section.

Two key assumptions of this model are worth examining. 
These set the stage for the subsequent developments in birth-death modelling of the bedload flux which I review next. 
First, this model assumes that stationary bed particles are in infinite supply. 
This is an implicit assumption: Nothing prevents the system from entraining particles ad infinitum through the entrainment processes characterized by rates $\lambda_0$ and $\mu$. 
Second, the model considers sediment transport in an isolated control volume, so the sediment flux is described probabilistically in a locality, but spatial correlations cannot be accounted for: the flux upstream of the control volume is also a fluctuating quantity, implying spatial correlations through immigration events into the control volume considered in this model.

These two observations support two of the more recent developments to the \citet{Ancey2008} birth-death immigration-emigration model: \citet{Turowski2009} generalized this model to finite bed material in order to describe transport within semi-alluvial bedrock environments, where alluvial deposits of limited quantity sit on top of bedrock. When all available particles are entrained and moved away from the control volume, there are none left to entrain, so the entrainment process is turned off. 
Likewise, \citet{Ancey2010, Ancey2014,Ancey2015} extended the birth-death process within a control volume as I reviewed here to an array of spatially distributed control volumes (cells). 
They showed that immigration and emigration processes between cells set up a bedload diffusion, and they connected this theory with earlier sediment diffusion theories by deriving an Exner type equation of sediment diffusion. 
I will now review each of these developments in turn. 
\end{comment}

>>>>

We can consider the flow of probability up and down the ladder of states ($n=0,1,2,\dots$) which is induced by these transitions, entrainment (including the collective entrainment term), deposition, immigration, and emigration, using the formalism of more general Markov birth-death processes \citep{Cox1965, Pielou1977}. 
By the standard way of developing master equations \citep[e.g.][]{Cox1965}, we find an infinite system of equations for the probabilities of finding $n$ moving particles within the control volume: 
\begin{multline} P(n,t+\delta t) = \alpha \delta t (n+1)P(n+1,t) \\+  [\lambda + (n-1)\mu]\delta t P(n-1,t) \\ + [1-\delta t[\lambda + n\alpha + n\mu]]P(n,t). \label{eq:ancey2008master}\end{multline}
Here $\alpha = \nu + \sigma$ summarizes the contributions of emigration and deposition, which both act to decrease the number of particles $n$, and $\lambda = \lambda_0 + \lambda_1$ summarizes the contributions of immigration and (individual) entrainment, which both act to increase the number of particles $n$.

@@

In effect, the probability generating function exchanges the discrete variable $n$ for a continous variable $z$. 
This first order partial differential equation for $G$ can be solved by the method of characteristics \citep{Cox1965, Garabedian1964}. 
If there are initially $N_0$ moving particles within the control volume, the solution is 
\be G(z,t) = \Big(\frac{\alpha-\mu}{(K-1)\mu z + \alpha - K \mu}\Big)^{n+\lambda/\mu}\Big( \frac{(K\alpha-\mu)z + \alpha(1-K)}{\alpha-\mu}\Big)^n. \label{eq:generator}\ee
The factor $K$ is the autocorrelation function $K(t) = \exp(-t(\alpha-\mu)).$

@@

As discussed, the negative binomial distribution supports wide fluctuations, controlled by the collective entrainment parameter $\mu$. 
The mean number of particles in the control volume, taken over the distribution \ref{eq:negbin}, is 
\be \bra n \ket = \sum_{n=0}^\infty n P(n) = \frac{\lambda \alpha}{\alpha-\mu}. \label{eq:anceymean}\ee
Likewise the variance is 
\be \bra \delta n^2 \ket = \sum_{n=0}^\infty (n-\bra n\ket)^2 P(n) = \frac{ \lambda \alpha}{\alpha-\mu^2},\ee
which means their ratio is 
\be \frac{\bra \delta n^2\ket }{\bra n\ket} = \frac{1}{1-\mu/\alpha}.\ee
It is interesting to compare this to equation \ref{eq:2006flucts} bounding the magnitude of fluctuations in the \citet{Ancey2006} model. 
In contrast, within the \citet{Ancey2008} model, the variance must exceed the mean, because as discussed, in steady state $\alpha>\mu$. 
The magnitude of fluctuations can grow arbitrarily with the collective entrainment parameter $\mu$. 
This is a coherent conclusion, since collective entrainment was introduced in order to represent the collective effects of turbulent fluctuations, collision-induced entrainment, and granular avalanches; and with the expressed intent of generating realistically wide bedload fluctuations, exceeding those of the \citet{Ancey2006} model.

Armed with the probability distribution for the number of particles in the control volume, equation \ref{eq:negbin}, and the link between control volume and surface statistics derived in section \ref{sec:fluxdef}, equation \ref{eq:fluxy}, we can derive the probability distribution of the bedload flux in a similar way as equation \ref{eq:anceygauss}.
Using the law for transformation of probabilities gives 
\be P(q_s) \approx \frac{\Gamma(r+\frac{Lw}{\nu_p u_s} q_s)}{\Gamma(r)\Gamma(1+\frac{Lw}{\nu_p u_s} q_s)}p^r(1-p)^{\frac{Lw}{\nu_p u_s} q_s}\ee
as the probability distribution of the bedload flux $q_s$.
The bedload flux, as derived in \citet{Ancey2008}, is a random variable with a negative binomial distribution.

Now we link back to the \citet{Einstein1950} theory. 
Einstein considered neither immigration ($\lambda_0$), emigration ($\nu$), nor collective entrainment ($\mu$). 
When all of these parameters are set to zero, $\mu = \lambda_0 = \nu = 0$, the parameters $r$ and $p$ of the distribution \ref{eq:negbin0} become $r \rightarrow \infty$ and $p=0$. 
With these limits, the distribution \ref{eq:negbin0} tends to a Poisson distribution. 
There are several ways to show this. 
The simplest way is to take the appropriate limits ($t \rightarrow \infty$, $\mu=\lambda_1=\nu= 0$)  into the generating function \ref{eq:generator}, giving
\be G(z) = e^{-\lambda_0(z-1)/\sigma}.\ee
This inverts to ($P(n) = \frac{1}{n!}(\partial/\partial z)^n G(z)|_{z=1}$)
\be P(n) = \frac{(\lambda_1/\sigma)^n}{n!}e^{-\lambda_1/\sigma},\label{eq:anceypoisson}\ee
the Poisson distribution with rate $\lambda_0/\sigma$.
The \citet{Ancey2008} model links to the limiting result of \citet{Ancey2006}, equation \ref{eq:ancey2006limit2}.

Using equation \ref{eq:fluxy} with this Poisson distribution, equation \ref{eq:anceypoisson}, we compute the number of active paricles $n$ in the absence of collective entrainment: 
\be \bra q_s \ket = \frac{\nu_p u_s}{L w} \sum_{n=0}^\infty n \frac{(\lambda_1/\sigma)^n}{n!}e^{-\lambda_1/\sigma} = \frac{\nu_p u_s}{L w} \frac{\lambda_1}{\sigma}. \label{eq:ancey2008mean}\ee
To connect equation \ref{eq:ancey2008mean} with the Einstein-like formula of Yalin, equation \ref{eq:yalinflux}, requires a delicate interpretation of the parameters $\lambda_0$ and $\sigma$. 
The rate of entrainment $\lambda_1$ pertains to all particles within the control volume, of which there is a very large number $N \rightarrow \infty$. 
Meanwhile, the rate of deposition $\sigma$ pertains to an individual particle. 
These interpretations are consistent with the master equation \ref{eq:ancey2008master}. 
Hence if we define the probability that (only) one particle entrains in a small interval $\delta t$ as $\lambda'$, then $\lambda = N\lambda'$, and the mean flux is 
\be \bra q_s \ket = \frac{\nu_p u_s}{L w}\frac{N\lambda'}{\sigma}.\ee
If we can again interpret the ratio $N/(Lw)$ as proportional to the area of an individual particle, using $\nu_p \propto a^3$, and the travel distance $l\propto a$, we get 
\be \bra q_s \ket = A \frac{a l}{a/u_s}\frac{\lambda'}{\sigma}, \ee
again in accord with the Yalin result \ref{eq:yalinflux}.

We have shown that the assumptions of \citet{Einstein1950} can be mixed with stochastic mathematics to develop a more coherent and physically-based theory of the bedload flux, as performed by \citet{Ancey2006}. 
Since the fluctuations predicted by the Gaussian bedload flux of \citet{Ancey2006} are not sufficiently large to describe experiments, we have mimiced \citet{Ancey2008} to include collective motion effects in order to express bedload fluctuations of sufficient magnitude from a birth-death model. 
Turning off the extra processes in the \citet{Ancey2008} model leads to a coherent connection back to a limiting result of \citet{Ancey2006}, and further to back to Yalin's revision of Einstein \citep{Yalin1972}.

However, there is obviously some ambiguity around the number of particles available for motion, which was denoted $N$ in previous sections in this review. 
Apparently, the \citet{Ancey2008} model considers the $N\rightarrow \infty$ limit, while the probability that any one particle is observed in motion goes to zero, similar to the Poisson limit of the \citet{Ancey2006} model. 
Of course, the number of particles available for motion is not unlimited. 
Over a purely alluvial bed composed to gravel, it must be limited by the number of particles on the bed surface. 
Within the idealization of \citet{Einstein1950}, this constraint was expressed by the equation $N a^2 \propto Lw$.

In the collective model of \citet{Ancey2008}, limiting the number of particles available for motion is more difficult. 
The most obvious way would be to consider that the entrainment rate of all particles is proportional to the entrainment rate of a single particle, times the number of particles at rest, such as $\lambda'(N-n)$. 
However, this does not work because particles can also arrive from upstream, meaning $n$ can exceed $N$. 
In this case, to consider the rate of entrainment decreases with the number of active particles, one would need to disciminate how the particles got into the control volume in the first place. 
Did they entrain, or did they emigrate?
This is non-Markovian. 
A second way to impose that a finite number of bed particles are available for entrainment is to consider the Markovian dynamics of two species at once: the bed surface particles, and the particles in motion. 
This path was taken by \citet{Turowski2009}, and this is the next section of the review.
